---
title: "Players"
execute:
  echo: false
  warning: false
---

```{r}
#| message: false
library(tidyverse)

# Load Mike Trout pitch & zone swing data (2014 vs. 2019)
trout_zone <- read_csv("data/trout_zone_swing.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 MVP)", "Post-Statcast (2019 MVP)"))
trout_pitches <- read_csv("data/trout_pitches.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 MVP)", "Post-Statcast (2019 MVP)"))

# Load Paul Goldschmidt pitch & zone swing data (2013 vs. 2022)
goldy_zone <- read_csv("data/goldschmidt_zone_swing.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2013 MVP Runner-Up)", "Post-Statcast (2022 MVP)"))
goldy_pitches <- read_csv("data/goldschmidt_pitches.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2013 MVP Runner-Up)", "Post-Statcast (2022 MVP)"))

# Load J.D. Martinez pitch & zone swing data (2014 vs. 2018)
martinez_zone <- read_csv("data/martinez_zone_swing.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 Breakout)", "Post-Statcast (2018 All-Star)"))
martinez_pitches <- read_csv("data/martinez_pitches.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 Breakout)", "Post-Statcast (2018 All-Star)"))

# Load Batting Average by Zone datasets
trout_zone_ba <- read_csv("data/trout_zone_ba.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 MVP)", "Post-Statcast (2019 MVP)"))
trout_at_bats <- read_csv("data/trout_at_bats.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 MVP)", "Post-Statcast (2019 MVP)"))

goldy_zone_ba <- read_csv("data/goldschmidt_zone_ba.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2013 MVP Runner-Up)", "Post-Statcast (2022 MVP)"))
goldy_at_bats <- read_csv("data/goldschmidt_at_bats.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2013 MVP Runner-Up)", "Post-Statcast (2022 MVP)"))

martinez_zone_ba <- read_csv("data/martinez_zone_ba.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 Breakout)", "Post-Statcast (2018 All-Star)"))
martinez_at_bats <- read_csv("data/martinez_at_bats.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 Breakout)", "Post-Statcast (2018 All-Star)"))
```

::: {.hero-banner}
# Player Strike Zone Case Studies
How did elite hitters respond to the Statcast era? By examining pitch-by-pitch tracking data across pre- and post-Statcast seasons, we trace how superstars like Mike Trout, Paul Goldschmidt, and J.D. Martinez adapted their swing decisions and zone coverage.
:::

::: {.panel-tabset}

## Mike Trout (2014 vs. 2019 MVP)

Analyzing Mike Trout's strike zone between 2014 and 2019 reveals how pitch-tracking data refined modern plate discipline and run production. Hitters learned to become hyper-selective—focusing swings on pitches in the heart of the zone where they do maximum damage, while laying off pitcher-friendly edge locations. Comparing his swing decisions directly with his batting average highlights where his selectivity paid off.

```{r}
#| label: plot-trout-strikezone
#| fig-width: 8.5
#| fig-height: 4.8
ggplot() +
  # Semi-transparent zone tiles colored by swing rate
  geom_tile(
    data = trout_zone,
    aes(x = factor(x), y = factor(y), fill = swing_pct),
    alpha = 0.35,
    color = "grey40",
    linewidth = 1.2
  ) +
  # Individual pitch dots jittered inside their respective zone
  geom_jitter(
    data = trout_pitches,
    aes(x = factor(x), y = factor(y), color = outcome),
    width = 0.35,
    height = 0.35,
    alpha = 0.65,
    size = 1.8
  ) +
  # Zone swing % labels
  geom_label(
    data = trout_zone,
    aes(x = factor(x), y = factor(y), label = paste0(round(swing_pct, 1), "%")),
    alpha = 0.75,
    fontface = "bold",
    size = 3.2,
    color = "black"
  ) +
  scale_fill_gradient(low = "#4575b4", high = "#d73027", name = "Zone Swing %") +
  scale_color_manual(values = c("Take (Called Strike)" = "#7f7f7f", "Swing" = "#d73027"), name = "Pitch Decision") +
  scale_x_discrete(labels = c("Inside", "Middle", "Away")) +
  scale_y_discrete(labels = c("Low", "Middle", "High")) +
  facet_wrap(~era) +
  coord_fixed() +
  labs(
    title = "Mike Trout Strike Zone: Pitch Decisions & Swing Rates",
    subtitle = "Points represent in-zone pitches (grey = take, red = swing) over zone swing %",
    x = "Horizontal Location (to RHB)",
    y = "Vertical Location"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 13)
  )
```

```{r}
#| label: plot-trout-ba
#| fig-width: 8.5
#| fig-height: 4.8
ggplot() +
  # Semi-transparent zone tiles colored by batting average
  geom_tile(
    data = trout_zone_ba,
    aes(x = factor(x), y = factor(y), fill = ba),
    alpha = 0.35,
    color = "grey40",
    linewidth = 1.2
  ) +
  # Individual at-bat dots jittered inside their respective zone
  geom_jitter(
    data = trout_at_bats,
    aes(x = factor(x), y = factor(y), color = outcome),
    width = 0.35,
    height = 0.35,
    alpha = 0.65,
    size = 1.8
  ) +
  # Zone BA labels
  geom_label(
    data = trout_zone_ba,
    aes(x = factor(x), y = factor(y), label = ba_label),
    alpha = 0.75,
    fontface = "bold",
    size = 3.2,
    color = "black"
  ) +
  scale_fill_gradient(low = "#4575b4", high = "#d73027", name = "Zone BA", limits = c(0, 0.55)) +
  scale_color_manual(values = c("Out" = "#7f7f7f", "Hit" = "#d73027"), name = "AB Outcome") +
  scale_x_discrete(labels = c("Inside", "Middle", "Away")) +
  scale_y_discrete(labels = c("Low", "Middle", "High")) +
  facet_wrap(~era) +
  coord_fixed() +
  labs(
    title = "Mike Trout Strike Zone: Batting Average by Location",
    subtitle = "Points represent individual at-bats (grey = out, red = hit) over zone batting average",
    x = "Horizontal Location (to RHB)",
    y = "Vertical Location"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 13)
  )
```

Between 2014 and 2019, Trout dialed his aggression into the middle of the plate—increasing his swing rate in the heart of the zone from **62.7%** to **77.4%**—while being far more patient on elevated inside pitches (dropping from **54.5%** to **36.8%**). This selectivity mirrored his results: in 2014, Trout was lethal in the lower and middle regions of the zone (hitting **.456** low-middle, **.440** low-inside, and **.431** in the heart), but struggled on elevated pitches, batting just **.053** up-middle. By 2019, Trout dramatically improved his coverage of high pitches—raising his up-middle batting average to **.257**—while hitting **.435** on pitches low-away.

## Paul Goldschmidt (Pre- vs. Post-Statcast MVP)

Paul Goldschmidt provides a compelling look at how an elite slugger adapted his swing decisions across nearly a decade of pitch tracking. In his 2013 breakout season with the Arizona Diamondbacks (where he was the NL MVP runner-up), Goldschmidt was particularly eager to attack pitches elevated in the zone, recording a **76.7% swing rate** in the upper-middle quadrant and **71.1%** mid-in. By his 2022 NL MVP campaign with the St. Louis Cardinals, Goldschmidt adopted a more balanced approach—maintaining a **66.8% swing rate** in the heart of the zone while becoming more selective on low and away pitches (**49.5%** swing rate).

```{r}
#| label: plot-goldschmidt-strikezone
#| fig-width: 8.5
#| fig-height: 4.8
ggplot() +
  # Semi-transparent zone tiles colored by swing rate
  geom_tile(
    data = goldy_zone,
    aes(x = factor(x), y = factor(y), fill = swing_pct),
    alpha = 0.35,
    color = "grey40",
    linewidth = 1.2
  ) +
  # Individual pitch dots jittered inside their respective zone
  geom_jitter(
    data = goldy_pitches,
    aes(x = factor(x), y = factor(y), color = outcome),
    width = 0.35,
    height = 0.35,
    alpha = 0.65,
    size = 1.8
  ) +
  # Zone swing % labels
  geom_label(
    data = goldy_zone,
    aes(x = factor(x), y = factor(y), label = paste0(round(swing_pct, 1), "%")),
    alpha = 0.75,
    fontface = "bold",
    size = 3.2,
    color = "black"
  ) +
  scale_fill_gradient(low = "#4575b4", high = "#d73027", name = "Zone Swing %") +
  scale_color_manual(values = c("Take (Called Strike)" = "#7f7f7f", "Swing" = "#d73027"), name = "Pitch Decision") +
  scale_x_discrete(labels = c("Inside", "Middle", "Away")) +
  scale_y_discrete(labels = c("Low", "Middle", "High")) +
  facet_wrap(~era) +
  coord_fixed() +
  labs(
    title = "Paul Goldschmidt Strike Zone: Pitch Decisions & Swing Rates",
    subtitle = "Points represent in-zone pitches (grey = take, red = swing) over zone swing %",
    x = "Horizontal Location (to RHB)",
    y = "Vertical Location"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 13)
  )
```

```{r}
#| label: plot-goldschmidt-ba
#| fig-width: 8.5
#| fig-height: 4.8
ggplot() +
  # Semi-transparent zone tiles colored by batting average
  geom_tile(
    data = goldy_zone_ba,
    aes(x = factor(x), y = factor(y), fill = ba),
    alpha = 0.35,
    color = "grey40",
    linewidth = 1.2
  ) +
  # Individual at-bat dots jittered inside their respective zone
  geom_jitter(
    data = goldy_at_bats,
    aes(x = factor(x), y = factor(y), color = outcome),
    width = 0.35,
    height = 0.35,
    alpha = 0.65,
    size = 1.8
  ) +
  # Zone BA labels
  geom_label(
    data = goldy_zone_ba,
    aes(x = factor(x), y = factor(y), label = ba_label),
    alpha = 0.75,
    fontface = "bold",
    size = 3.2,
    color = "black"
  ) +
  scale_fill_gradient(low = "#4575b4", high = "#d73027", name = "Zone BA", limits = c(0, 0.55)) +
  scale_color_manual(values = c("Out" = "#7f7f7f", "Hit" = "#d73027"), name = "AB Outcome") +
  scale_x_discrete(labels = c("Inside", "Middle", "Away")) +
  scale_y_discrete(labels = c("Low", "Middle", "High")) +
  facet_wrap(~era) +
  coord_fixed() +
  labs(
    title = "Paul Goldschmidt Strike Zone: Batting Average by Location",
    subtitle = "Points represent individual at-bats (grey = out, red = hit) over zone batting average",
    x = "Horizontal Location (to RHB)",
    y = "Vertical Location"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 13)
  )
```

Comparing his 2013 and 2022 seasons reveals how Goldschmidt developed a patient, disciplined approach across the strike zone—remaining aggressive in the heart (**66.8%**) while cutting swings on outer-edge chase zones. Examining his batting average highlights his shift from a high-ball hitter in Arizona to an all-fields, lower-zone punisher in St. Louis. In 2013, Goldschmidt crushed pitches in the upper third of the strike zone, hitting **.432** up-and-away and **.405** up-middle. In 2022, pitchers combated his high-ball power, forcing Goldschmidt to adjust down—where he batted **.397** in the heart, **.391** low-middle, and **.350** low-and-away.



## J.D. Martinez (Pre- vs. Post-Statcast Breakout)

After overhauling his swing mechanics prior to the 2014 season, J.D. Martinez emerged as one of baseball's premier sluggers. By his 2018 championship season with the Boston Red Sox, Statcast data and launch angle optimization helped him refine his plate aggression even further.

```{r}
#| label: plot-martinez-strikezone
#| fig-width: 8.5
#| fig-height: 4.8
ggplot() +
  # Semi-transparent zone tiles colored by swing rate
  geom_tile(
    data = martinez_zone,
    aes(x = factor(x), y = factor(y), fill = swing_pct),
    alpha = 0.35,
    color = "grey40",
    linewidth = 1.2
  ) +
  # Individual pitch dots jittered inside their respective zone
  geom_jitter(
    data = martinez_pitches,
    aes(x = factor(x), y = factor(y), color = outcome),
    width = 0.35,
    height = 0.35,
    alpha = 0.65,
    size = 1.8
  ) +
  # Zone swing % labels
  geom_label(
    data = martinez_zone,
    aes(x = factor(x), y = factor(y), label = paste0(round(swing_pct, 1), "%")),
    alpha = 0.75,
    fontface = "bold",
    size = 3.2,
    color = "black"
  ) +
  scale_fill_gradient(low = "#4575b4", high = "#d73027", name = "Zone Swing %") +
  scale_color_manual(values = c("Take (Called Strike)" = "#7f7f7f", "Swing" = "#d73027"), name = "Pitch Decision") +
  scale_x_discrete(labels = c("Inside", "Middle", "Away")) +
  scale_y_discrete(labels = c("Low", "Middle", "High")) +
  facet_wrap(~era) +
  coord_fixed() +
  labs(
    title = "J.D. Martinez Strike Zone: Pitch Decisions & Swing Rates",
    subtitle = "Points represent in-zone pitches (grey = take, red = swing) over zone swing %",
    x = "Horizontal Location (to RHB)",
    y = "Vertical Location"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 13)
  )
```

```{r}
#| label: plot-martinez-ba
#| fig-width: 8.5
#| fig-height: 4.8
ggplot() +
  # Semi-transparent zone tiles colored by batting average
  geom_tile(
    data = martinez_zone_ba,
    aes(x = factor(x), y = factor(y), fill = ba),
    alpha = 0.35,
    color = "grey40",
    linewidth = 1.2
  ) +
  # Individual at-bat dots jittered inside their respective zone
  geom_jitter(
    data = martinez_at_bats,
    aes(x = factor(x), y = factor(y), color = outcome),
    width = 0.35,
    height = 0.35,
    alpha = 0.65,
    size = 1.8
  ) +
  # Zone BA labels
  geom_label(
    data = martinez_zone_ba,
    aes(x = factor(x), y = factor(y), label = ba_label),
    alpha = 0.75,
    fontface = "bold",
    size = 3.2,
    color = "black"
  ) +
  scale_fill_gradient(low = "#4575b4", high = "#d73027", name = "Zone BA", limits = c(0, 0.55)) +
  scale_color_manual(values = c("Out" = "#7f7f7f", "Hit" = "#d73027"), name = "AB Outcome") +
  scale_x_discrete(labels = c("Inside", "Middle", "Away")) +
  scale_y_discrete(labels = c("Low", "Middle", "High")) +
  facet_wrap(~era) +
  coord_fixed() +
  labs(
    title = "J.D. Martinez Strike Zone: Batting Average by Location",
    subtitle = "Points represent individual at-bats (grey = out, red = hit) over zone batting average",
    x = "Horizontal Location (to RHB)",
    y = "Vertical Location"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 13)
  )
```

Comparing Martinez's 2014 breakout in Detroit with his 2018 championship season in Boston highlights how his refined plate discipline fueled elite production. Martinez became markedly more aggressive on pitches elevated in the zone—increasing his swing rate up-and-in from **73.9%** to **82.6%**—while demonstrating greater patience low-and-in (dropping from **78.6%** to **68.3%**). Most dramatically, Statcast-driven launch angle optimization helped him expand his damage zone to the outer half: his batting average on pitches up-and-away surged from **.000** in 2014 to **.526** in 2018, alongside a **.517** mark middle-inside.

:::


