---
title: "Players"
execute:
  echo: false
  warning: false
---

```{r}
#| message: false
library(tidyverse)

# Load Mike Trout pitch & zone swing data (2014 vs. 2019)
trout_zone <- read_csv("data/trout_zone_swing.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 MVP)", "Post-Statcast (2019 MVP)"))
trout_pitches <- read_csv("data/trout_pitches.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 MVP)", "Post-Statcast (2019 MVP)"))

# Load Paul Goldschmidt pitch & zone swing data (2013 vs. 2022)
goldy_zone <- read_csv("data/goldschmidt_zone_swing.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2013 MVP Runner-Up)", "Post-Statcast (2022 MVP)"))
goldy_pitches <- read_csv("data/goldschmidt_pitches.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2013 MVP Runner-Up)", "Post-Statcast (2022 MVP)"))

# Load Giancarlo Stanton pitch & zone swing data (2014 vs. 2017)
stanton_zone <- read_csv("data/stanton_zone_swing.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 MVP Runner-Up)", "Post-Statcast (2017 MVP)"))
stanton_pitches <- read_csv("data/stanton_pitches.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 MVP Runner-Up)", "Post-Statcast (2017 MVP)"))

# Load Batting Average by Zone datasets
trout_zone_ba <- read_csv("data/trout_zone_ba.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 MVP)", "Post-Statcast (2019 MVP)"))
trout_at_bats <- read_csv("data/trout_at_bats.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 MVP)", "Post-Statcast (2019 MVP)"))

goldy_zone_ba <- read_csv("data/goldschmidt_zone_ba.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2013 MVP Runner-Up)", "Post-Statcast (2022 MVP)"))
goldy_at_bats <- read_csv("data/goldschmidt_at_bats.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2013 MVP Runner-Up)", "Post-Statcast (2022 MVP)"))

stanton_zone_ba <- read_csv("data/stanton_zone_ba.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 MVP Runner-Up)", "Post-Statcast (2017 MVP)"))
stanton_at_bats <- read_csv("data/stanton_at_bats.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 MVP Runner-Up)", "Post-Statcast (2017 MVP)"))
```

## Strike Zone Swing Rate: Mike Trout (Pre- vs. Post-Statcast MVP)

Analyzing Mike Trout's strike zone swing rates between 2014 and 2019 reveals how granular tracking data refined modern plate discipline. As pitch-tracking systems like Statcast provided hitters with detailed heat maps of their run production, hitters learned to become hyper-selective—focusing swings on pitches in the heart of the zone where they do maximum damage, while laying off pitcher-friendly edge locations. In 2014, Trout was relatively balanced across the zone, swinging at **62.7%** of pitches over the middle and **54.5%** in the upper-inside corner. By his 2019 MVP season, his swing rate in the heart of the plate climbed to **77.4%**, while he cut down his swings in the upper-inside quadrant to just **36.8%**.

```{r}
#| label: plot-trout-strikezone
#| fig-width: 8.5
#| fig-height: 4.8
ggplot() +
  # Semi-transparent zone tiles colored by swing rate
  geom_tile(
    data = trout_zone,
    aes(x = factor(x), y = factor(y), fill = swing_pct),
    alpha = 0.35,
    color = "grey40",
    linewidth = 1.2
  ) +
  # Individual pitch dots jittered inside their respective zone
  geom_jitter(
    data = trout_pitches,
    aes(x = factor(x), y = factor(y), color = outcome),
    width = 0.35,
    height = 0.35,
    alpha = 0.65,
    size = 1.8
  ) +
  # Zone swing % labels
  geom_label(
    data = trout_zone,
    aes(x = factor(x), y = factor(y), label = paste0(round(swing_pct, 1), "%")),
    alpha = 0.75,
    fontface = "bold",
    size = 3.2,
    color = "black"
  ) +
  scale_fill_gradient(low = "#4575b4", high = "#d73027", name = "Zone Swing %") +
  scale_color_manual(values = c("Take (Called Strike)" = "#7f7f7f", "Swing" = "#d73027"), name = "Pitch Decision") +
  scale_x_discrete(labels = c("Inside", "Middle", "Away")) +
  scale_y_discrete(labels = c("Low", "Middle", "High")) +
  facet_wrap(~era) +
  coord_fixed() +
  labs(
    title = "Mike Trout Strike Zone: Pitch Decisions & Swing Rates",
    subtitle = "Points represent in-zone pitches (grey = take, red = swing) over zone swing %",
    x = "Horizontal Location (to RHB)",
    y = "Vertical Location"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 13)
  )
```

Between 2014 and 2019, Trout dialed his aggression into the middle of the plate—increasing his swing rate in the heart of the zone from **62.7%** to **77.4%**—while being far more patient on elevated inside pitches (dropping from **54.5%** to **36.8%**).

### Mike Trout: Batting Average by Strike Zone Location

While swing rates illustrate where a hitter chooses to swing, batting average by zone reveals where those swings actually produce hits. In 2014, Trout was lethal in the lower and middle regions of the zone (hitting **.456** low-middle, **.440** low-inside, and **.431** in the heart), but struggled on elevated pitches, batting just **.053** up-middle. By 2019, Trout dramatically improved his coverage of high pitches—raising his up-middle batting average to **.257**—while hitting **.435** on pitches low-away.

```{r}
#| label: plot-trout-ba
#| fig-width: 8.5
#| fig-height: 4.8
ggplot() +
  # Semi-transparent zone tiles colored by batting average
  geom_tile(
    data = trout_zone_ba,
    aes(x = factor(x), y = factor(y), fill = ba),
    alpha = 0.35,
    color = "grey40",
    linewidth = 1.2
  ) +
  # Individual at-bat dots jittered inside their respective zone
  geom_jitter(
    data = trout_at_bats,
    aes(x = factor(x), y = factor(y), color = outcome),
    width = 0.35,
    height = 0.35,
    alpha = 0.65,
    size = 1.8
  ) +
  # Zone BA labels
  geom_label(
    data = trout_zone_ba,
    aes(x = factor(x), y = factor(y), label = ba_label),
    alpha = 0.75,
    fontface = "bold",
    size = 3.2,
    color = "black"
  ) +
  scale_fill_gradient(low = "#4575b4", high = "#d73027", name = "Zone BA", limits = c(0, 0.55)) +
  scale_color_manual(values = c("Out" = "#7f7f7f", "Hit" = "#d73027"), name = "AB Outcome") +
  scale_x_discrete(labels = c("Inside", "Middle", "Away")) +
  scale_y_discrete(labels = c("Low", "Middle", "High")) +
  facet_wrap(~era) +
  coord_fixed() +
  labs(
    title = "Mike Trout Strike Zone: Batting Average by Location",
    subtitle = "Points represent individual at-bats (grey = out, red = hit) over zone batting average",
    x = "Horizontal Location (to RHB)",
    y = "Vertical Location"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 13)
  )
```

## Strike Zone Swing Rate: Paul Goldschmidt (Pre- vs. Post-Statcast MVP)

Paul Goldschmidt provides a compelling look at how an elite slugger adapted his swing decisions across nearly a decade of pitch tracking. In his 2013 breakout season with the Arizona Diamondbacks (where he was the NL MVP runner-up), Goldschmidt was particularly eager to attack pitches elevated in the zone, recording a **76.7% swing rate** in the upper-middle quadrant and **71.1%** mid-in. By his 2022 NL MVP campaign with the St. Louis Cardinals, Goldschmidt adopted a more balanced approach—maintaining a **66.8% swing rate** in the heart of the zone while becoming more selective on low and away pitches (**49.5%** swing rate).

```{r}
#| label: plot-goldschmidt-strikezone
#| fig-width: 8.5
#| fig-height: 4.8
ggplot() +
  # Semi-transparent zone tiles colored by swing rate
  geom_tile(
    data = goldy_zone,
    aes(x = factor(x), y = factor(y), fill = swing_pct),
    alpha = 0.35,
    color = "grey40",
    linewidth = 1.2
  ) +
  # Individual pitch dots jittered inside their respective zone
  geom_jitter(
    data = goldy_pitches,
    aes(x = factor(x), y = factor(y), color = outcome),
    width = 0.35,
    height = 0.35,
    alpha = 0.65,
    size = 1.8
  ) +
  # Zone swing % labels
  geom_label(
    data = goldy_zone,
    aes(x = factor(x), y = factor(y), label = paste0(round(swing_pct, 1), "%")),
    alpha = 0.75,
    fontface = "bold",
    size = 3.2,
    color = "black"
  ) +
  scale_fill_gradient(low = "#4575b4", high = "#d73027", name = "Zone Swing %") +
  scale_color_manual(values = c("Take (Called Strike)" = "#7f7f7f", "Swing" = "#d73027"), name = "Pitch Decision") +
  scale_x_discrete(labels = c("Inside", "Middle", "Away")) +
  scale_y_discrete(labels = c("Low", "Middle", "High")) +
  facet_wrap(~era) +
  coord_fixed() +
  labs(
    title = "Paul Goldschmidt Strike Zone: Pitch Decisions & Swing Rates",
    subtitle = "Points represent in-zone pitches (grey = take, red = swing) over zone swing %",
    x = "Horizontal Location (to RHB)",
    y = "Vertical Location"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 13)
  )
```

Comparing his 2013 and 2022 seasons reveals how Goldschmidt developed a patient, disciplined approach across the strike zone—remaining aggressive in the heart (**66.8%**) while cutting swings on outer-edge chase zones.

### Paul Goldschmidt: Batting Average by Strike Zone Location

Examining Goldschmidt's batting average by zone highlights his shift from a high-ball hitter in Arizona to an all-fields, lower-zone punisher in St. Louis. In his 2013 season, Goldschmidt crushed pitches in the upper third of the strike zone, hitting **.432** up-and-away and **.405** up-middle. In his 2022 NL MVP campaign, pitchers combated his high-ball power, forcing Goldschmidt to adjust down—where he batted **.397** in the heart, **.391** low-middle, and **.350** low-and-away.

```{r}
#| label: plot-goldschmidt-ba
#| fig-width: 8.5
#| fig-height: 4.8
ggplot() +
  # Semi-transparent zone tiles colored by batting average
  geom_tile(
    data = goldy_zone_ba,
    aes(x = factor(x), y = factor(y), fill = ba),
    alpha = 0.35,
    color = "grey40",
    linewidth = 1.2
  ) +
  # Individual at-bat dots jittered inside their respective zone
  geom_jitter(
    data = goldy_at_bats,
    aes(x = factor(x), y = factor(y), color = outcome),
    width = 0.35,
    height = 0.35,
    alpha = 0.65,
    size = 1.8
  ) +
  # Zone BA labels
  geom_label(
    data = goldy_zone_ba,
    aes(x = factor(x), y = factor(y), label = ba_label),
    alpha = 0.75,
    fontface = "bold",
    size = 3.2,
    color = "black"
  ) +
  scale_fill_gradient(low = "#4575b4", high = "#d73027", name = "Zone BA", limits = c(0, 0.55)) +
  scale_color_manual(values = c("Out" = "#7f7f7f", "Hit" = "#d73027"), name = "AB Outcome") +
  scale_x_discrete(labels = c("Inside", "Middle", "Away")) +
  scale_y_discrete(labels = c("Low", "Middle", "High")) +
  facet_wrap(~era) +
  coord_fixed() +
  labs(
    title = "Paul Goldschmidt Strike Zone: Batting Average by Location",
    subtitle = "Points represent individual at-bats (grey = out, red = hit) over zone batting average",
    x = "Horizontal Location (to RHB)",
    y = "Vertical Location"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 13)
  )
```

## Strike Zone Swing Rate: Giancarlo Stanton (Pre- vs. Post-Statcast MVP)

Giancarlo Stanton's historical power in his 59-home-run 2017 MVP season highlights his willingness to do damage on hittable pitches across the zone. In 2014, when Stanton finished as the NL MVP runner-up, he attacked pitches inside and over the heart aggressively (**72.3%** in the center and **75.4%** up-and-in), while swinging on **56.9%** of low-away pitches. By 2017, Stanton became even more aggressive on pitches up in the zone, swinging at an overwhelming **79.7%** of up-and-in pitches and **75.4%** in the heart, while laying off pitches low-and-away (**52.2%** swing rate).

```{r}
#| label: plot-stanton-strikezone
#| fig-width: 8.5
#| fig-height: 4.8
ggplot() +
  # Semi-transparent zone tiles colored by swing rate
  geom_tile(
    data = stanton_zone,
    aes(x = factor(x), y = factor(y), fill = swing_pct),
    alpha = 0.35,
    color = "grey40",
    linewidth = 1.2
  ) +
  # Individual pitch dots jittered inside their respective zone
  geom_jitter(
    data = stanton_pitches,
    aes(x = factor(x), y = factor(y), color = outcome),
    width = 0.35,
    height = 0.35,
    alpha = 0.65,
    size = 1.8
  ) +
  # Zone swing % labels
  geom_label(
    data = stanton_zone,
    aes(x = factor(x), y = factor(y), label = paste0(round(swing_pct, 1), "%")),
    alpha = 0.75,
    fontface = "bold",
    size = 3.2,
    color = "black"
  ) +
  scale_fill_gradient(low = "#4575b4", high = "#d73027", name = "Zone Swing %") +
  scale_color_manual(values = c("Take (Called Strike)" = "#7f7f7f", "Swing" = "#d73027"), name = "Pitch Decision") +
  scale_x_discrete(labels = c("Inside", "Middle", "Away")) +
  scale_y_discrete(labels = c("Low", "Middle", "High")) +
  facet_wrap(~era) +
  coord_fixed() +
  labs(
    title = "Giancarlo Stanton Strike Zone: Pitch Decisions & Swing Rates",
    subtitle = "Points represent in-zone pitches (grey = take, red = swing) over zone swing %",
    x = "Horizontal Location (to RHB)",
    y = "Vertical Location"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 13)
  )
```

Comparing Stanton's 2014 and 2017 campaigns shows his aggressive upper-zone coverage—driving his up-and-in swing rate close to **80%** while cutting down on chase swings low and away.

### Giancarlo Stanton: Batting Average by Strike Zone Location

Looking at Stanton's batting average by zone illustrates how pitchers attacked him differently between his two MVP-caliber seasons. In 2014, Stanton punished mistake pitches across the upper quadrant, batting **.538** up-and-away and **.400** up-and-in, alongside a **.462** mark in the heart. In his 2017 MVP season, pitchers combated his power with elevated velocity at the top of the zone (holding him to **.200** up-middle and **.217** up-in). Stanton compensated by doing immense damage on pitches down in the zone—batting **.383** low-inside and **.371** low-middle while crushing **.410** in the heart.

```{r}
#| label: plot-stanton-ba
#| fig-width: 8.5
#| fig-height: 4.8
ggplot() +
  # Semi-transparent zone tiles colored by batting average
  geom_tile(
    data = stanton_zone_ba,
    aes(x = factor(x), y = factor(y), fill = ba),
    alpha = 0.35,
    color = "grey40",
    linewidth = 1.2
  ) +
  # Individual at-bat dots jittered inside their respective zone
  geom_jitter(
    data = stanton_at_bats,
    aes(x = factor(x), y = factor(y), color = outcome),
    width = 0.35,
    height = 0.35,
    alpha = 0.65,
    size = 1.8
  ) +
  # Zone BA labels
  geom_label(
    data = stanton_zone_ba,
    aes(x = factor(x), y = factor(y), label = ba_label),
    alpha = 0.75,
    fontface = "bold",
    size = 3.2,
    color = "black"
  ) +
  scale_fill_gradient(low = "#4575b4", high = "#d73027", name = "Zone BA", limits = c(0, 0.55)) +
  scale_color_manual(values = c("Out" = "#7f7f7f", "Hit" = "#d73027"), name = "AB Outcome") +
  scale_x_discrete(labels = c("Inside", "Middle", "Away")) +
  scale_y_discrete(labels = c("Low", "Middle", "High")) +
  facet_wrap(~era) +
  coord_fixed() +
  labs(
    title = "Giancarlo Stanton Strike Zone: Batting Average by Location",
    subtitle = "Points represent individual at-bats (grey = out, red = hit) over zone batting average",
    x = "Horizontal Location (to RHB)",
    y = "Vertical Location"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 13)
  )
```


