---
title: "Story"
execute:
  echo: false
  warning: false
---

```{r}
#| message: false
library(tidyverse)
library(plotly)

# Load player-season datasets to calculate annual averages
hitters_hard_hit <- read_csv("data/all_hitters_hard_hit.csv")
pitcher_seasons <- read_csv("data/pitcher_seasons.csv")

yearly_hh <- hitters_hard_hit |>
  group_by(year) |>
  summarize(avg_hard_hit = round(mean(hard_hit_pct, na.rm = TRUE), 1))

yearly_k9 <- pitcher_seasons |>
  group_by(yearID) |>
  summarize(avg_so9 = round(mean(so9), 2))

combined_trajectory <- inner_join(yearly_hh, yearly_k9, by = c("year" = "yearID")) |>
  mutate(era = if_else(year < 2015, "Pre-Statcast (2005–2014)", "Post-Statcast (2015–2025)"))
```

```{r}
#| label: plot-game-trajectory
plot_ly(data = combined_trajectory, x = ~avg_so9, y = ~avg_hard_hit, height = 450) |>
  # Chronological path connecting seasons
  add_paths(
    line = list(color = "#888888", width = 2, dash = "solid"),
    showlegend = FALSE,
    hoverinfo = "none"
  ) |>
  # Season marker points colored by era
  add_markers(
    color = ~era,
    colors = c("Pre-Statcast (2005–2014)" = "#377eb8", "Post-Statcast (2015–2025)" = "#ba122b"),
    marker = list(size = 9, line = list(color = "black", width = 1)),
    text = ~paste0("<b>Season: ", year, "</b><br>Era: ", era, "<br>Strikeouts per 9: ", avg_so9, "<br>Hard-Hit Rate: ", avg_hard_hit, "%"),
    hoverinfo = "text"
  ) |>
  # Year labels for key seasons
  add_text(
    text = ~year,
    textposition = "top center",
    showlegend = FALSE,
    textfont = list(size = 9, color = "#333333")
  ) |>
  layout(
    title = list(
      text = "<b>The Shift in Modern Baseball (2005–2025)</b><br><sup>League Average K/9 vs. Hard-Hit Rate Trajectory</sup>",
      x = 0.5,
      xanchor = "center",
      font = list(size = 15)
    ),
    margin = list(t = 70, b = 60, l = 60, r = 50),
    xaxis = list(
      title = "League Average Strikeouts per 9 Innings (K/9)",
      dtick = 0.5,
      gridcolor = "#f0f0f0"
    ),
    yaxis = list(
      title = "League Average Hard-Hit Rate (%)",
      ticksuffix = "%",
      gridcolor = "#f0f0f0"
    ),
    annotations = list(
      list(
        x = 6.55, y = 30.2,
        text = "<b>2005 (Start)</b>",
        showarrow = TRUE,
        arrowhead = 2,
        ax = -45, ay = 25
      ),
      list(
        x = 8.61, y = 42.1,
        text = "<b>2025 (Latest)</b>",
        showarrow = TRUE,
        arrowhead = 2,
        ax = 45, ay = -25
      )
    ),
    legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.18)
  ) |>
  config(responsive = TRUE)
```

## Analysis: The Two-Way Technological Surge

Tracing Major League Baseball from 2005 (bottom left) to 2025 (top right) reveals an unmistakable diagonal migration: the game simultaneously produced significantly higher strikeout rates and higher exit velocities. Rather than giving an edge to only pitchers or only hitters, tracking technology elevated both disciplines at once. Pitchers leveraged spin tracking and release metrics to drive league average strikeouts from **6.55 K/9 in 2005** to **8.61 K/9 in 2025** (a 31.5% increase). Concurrently, hitters used launch angle and exit velocity analytics to optimize swing paths for hard contact, lifting average hard-hit rates from **30.2%** up to **42.1%**. This trajectory confirms that technology did not simply tilt the scale toward one side—it redefined the physical standards of the sport.
